Pseudogapped Mott criticality: Stretching Kondo Screening to Breaking Point

### 2025-05-09 TilingProjectDraft/manuscript.tex
polished manuscript and started adding endmatter

\documentclass[%
 reprint,
superscriptaddress,
groupedaddress,
%unsortedaddress,
%runinaddress,
%frontmatterverbose, 
%preprint,
%preprintnumbers,
%nofootinbib,
%nobibnotes,
%bibnotes,
 amsmath,amssymb,
 aps,
prl
%prb,
%rmp,
%prstab,
%prstper,
%floatfix,
]{revtex4-2}

@@

\title{Pseudogapped non-Fermi liquid phase arising from Kondo breakdown at the Mott transition}

@@

\author{Abhirup Mukherjee}
\author{Siddhartha Lal}
\affiliation{%
 Department of Physical Sciences, Indian Institute of Science Education and Research Kolkata, Nadia - 741246, India
}%

\date{\today}
\begin{abstract}
We present an auxiliary model approach that illuminates the centrality of the pseudogap phenomenon in the Mott transition of the Hubbard-Heisenberg model on the half-filled square lattice. Passage from Fermi liquid metal to Mott insulator involves two stages: the first involves a systematic unraveling of Kondo screening due to frustrating charge fluctuations in the conduction bath of the underlying lattice-embedded quantum impurity model. The unraveling destabilises the Fermi liquid, culminating in disconnection of the Fermi surface by gapping the antinodes. In the second stage, an emergent pseudogap phase arises from an effective two-channel Kondo problem in the underlying impurity model. The resulting momentum space Fermi arcs possess non-Fermi liquid excitations proximate to a critical Fermi surface. Upon approaching the transition, the nature of this exotic metal evolves with shrinking size of the arcs to a singular nodal non-Fermi liquid along with the onset of long-ranged real-space entanglement and spin-flip correlations. This reveals the Mott transition on the square lattice to be well beyond the paradigm of local quantum criticality.
\end{abstract}

@@

\paragraph*{Introduction.}
The origin of the pseudogap and strange metal phases of the cuprates continue to be hotly debated topics in the context of high-temperature superconductivity~\cite{keimer2015quantum,ProustTaillefer2019}. The anomalous pseudogap (PG) phase, observed most prominently in the cuprates, exhibits nodal-antinodal dichotomy with spectral weight being concentrated on Fermi arcs around the node~\cite{loeserKapitulnik1996,Norman1998,Hashimoto2014}. While several methods have observed a PG in the Hubbard model ~\cite{KyungKotliar2006,MacridinAzevedo2006,WuFerrero2018,anirbanmott2,HilleAndergassen2020}, there is no general consensus yet regarding various aspects of the PG phase, such as its relation to the Mott insulating and superconducting phases proximate to it, how it evolves from weak-coupling to strong-coupling, and whether the nodal-antinodal dichotomy is intrinsic to the model.

In order to investigate the precise mechanism of transition/crossover connecting the metallic phases with the Mott insulating phase, we present a new auxiliary model method to study correlated electronic systems, and apply it to the particle-hole symmetric Hubbard-Heisenberg model to investigate the origin and nature of its pseudogap phase and the subsequent transition into a Mott insulator. The method requires identifying an appropriate impurity model which can faithfully capture the physics of lattice model we are trying to solve. Passage to the bulk model is accomplished by applying lattice translation operators on the impurity model, which makes it possible to relate various computables of the auxiliary model to those of the lattice model.

Our approach differs from dynamical mean-field approximation based methods (specifically the cellular and cluster-based avatars) in the fact there no self-consistency condition that needs to be imposed - the lattice model is explicitly constructed from a {\it lattice-embedded} impurity model. While this makes the impurity model trickier to solve, it makes more transparent the underlying local physics arising from correlations, and provides increased momentum-space resolution which is crucial for studying PG physics. Indeed, an important conclusion of our work that results from this is that the Mott transition on the square lattice appears to be well beyond the paradigm of local quantum criticality, due to the onset of long-ranged real-space entanglement and spin-flip correlations close to the transition.

\paragraph*{A lattice-embedded impurity model.}
%Crucial to our method is the choice of an impurity auxiliary model that is able 
In order to capture important local aspects of electronic correlations in the lattice model,  
%Some of us have shown in a previous work
%~\cite{Mukherjee_2023} that various aspects of the Mott metal-insulator transition on the infinite dimensional Hubbard model on the Bethe lattice (as seen, for example, from dynamical mean-field theory~\cite{georges1996}) can be captured from the breakdown of the strongly coupled Kondo phase of a simple extension of the Anderson impurity model (eSIAM). 
%Building on this insight, 
we embed an extended Anderson impurity 
%spin-$1/2$ moment ${\bf S}_d$ 
into the lattice of a half-filled 2D conduction bath (with nearest-neighbour hopping). This builds on insight 
%of an earlier work
\cite{Mukherjee_2023} that various aspects of the Mott %metal-insulator 
transition on the %infinite-dimensional 
Hubbard model on the Bethe lattice in infinite dimensions (as seen %, for example, 
from 
%dynamical mean-field theory
DMFT~\cite{georges1996}) can be captured from the breakdown of the strongly coupled Kondo phase of a simple extension of the Anderson impurity model (eSIAM). 
%This lowers the \(s-\)wave symmetry of Kondo interactions (on, say, an infinite-dimensional Bethe lattice) down to the \(C_4\) symmetry of the 2D square lattice.

In capturing the Mott transition, we focus on the %large-$U$ 
limit of large local correlations within the impurity model, where the impurity-bath hybridisation occurs through an antiferromagnetic Kondo coupling ($J>0$). The model is described by the Hamiltonian
\begin{equation}\begin{aligned}\label{impurityModel}
	\mathcal{H}_\text{aux} = H_\text{cbath} + H_\text{imp-cbath} + H_\text{cbath-int}~,
\end{aligned}\end{equation}
where $H_\text{cbath} = -2t\sum_{{\bf k}}\left[\cos(ak_x) + \cos(ak_y)\right] c^\dagger_{{\bf k},\sigma}c_{{\bf k},\sigma}$ is the kinetic energy for conduction bath tight-binding electrons ($c^\dagger_{{\bf k},\sigma}$) on a half-filled square lattice, and the impurity spin-1/2 moment (${\bf S}_d$) interacts with four nearest neighbour sites (henceforth labelled as \textit{zeroth} sites with index $Z$) 
%in \(H_\text{imp-cbath}\) 
through local Kondo terms of uniform strength \(H_\text{imp-cbath} =\frac{1}{2} J\sum_{\sigma_1,\sigma_2}\sum_{Z} {\bf S}_d\cdot c^\dagger_{Z\sigma_1}{\boldsymbol \tau}_{\sigma_1,\sigma_2} c_{Z\sigma_2}\).
%is the impurity-bath Kondo interaction. 
%and \(H_\text{cbath-int} = -\frac{W}{2}\sum_{Z} \left(n_{Z\uparrow} - n_{Z\downarrow}\right)^2\). 
%is a local intra-bath correlation that 
%models minimally the correlated nature of the conduction bath. 
%The index \(Z\) sums over all four nearest-neighbour sites of the impurity site, henceforth referred to as bath {\it zeroth} sites). 
A minimal model for correlations within the conduction bath that frustrate the Kondo screening is given by the third term:
%$H_\text{cbath-int}$ 
%represents local correlations on the four bath zeroth sites, 
\(H_\text{cbath-int} = -\frac{W}{2}\sum_{Z} \left(n_{Z\uparrow} - n_{Z\downarrow}\right)^2\). 
%, again of uniform strength. 
%In the above definitions, \(c^\dagger_{d\sigma}\) is the impurity site creation operator (for spin \(\sigma \)), while \(c^\dagger_{{\bf k}\sigma}\) is the conduction bath creation operator (for momentum \({\bf k}\) and spin \(\sigma\)).
The 2D structure of the impurity model, and its distinction from the infinite-dimensional counterpart, becomes apparent upon Fourier transforming the Kondo coupling to ${\bf k}$-space (with lattice spacing set to unity):
%once we transform the Hamiltonian to momentum space:
\begin{equation}\begin{aligned}
	J_{{\bf k}, {\bf k}^\prime} = \frac{J}{2}&\left[\cos\left({\bf k}_x - {\bf k}^\prime_x\right) + \cos\left({\bf k}_y - {\bf k}^\prime_y\right)\right]~,
\end{aligned}\end{equation}
with $J>0$.
%where the subscripts \(x\) and \(y\) represent components of the momenta vector, and the lattice spacing has been set to unity. This form shows that The Kondo coupling 
Clearly, $J_{{\bf k}, {\bf k}^\prime}$ is 
%no longer in
dependent on the exchange of momentum, and encodes the $C_{4}$ lattice symmetry. 
%geometry. Throughout the work
%In all that follows, we set the lattice spacing to unity.

\paragraph*{Reconstructing the Hubbard-Heisenberg model.} %from the auxiliary model: Formal prescription of our method.}
We will now define the {\it tiling} (or periodisation) procedure by which we recreate the lattice model 
%by using instances of 
from the 
%auxilliary
impurity model Hamiltonian. 
%In order to identify a 
For a {\it unit cell} 
%for 
of the tiling, 
%procedure, 
we consider an impurity model \(\mathcal{H}_\text{aux}({\bf r}_d)\) that has the impurity site at a reference site \({\bf r}_d\) of our lattice. In order to create the bulk model, we translate the unit cell across all sites of the lattice:
\begin{equation}\begin{aligned}
	\label{tilingPrescription}
	\mathcal{H}_\text{tiled} &= \sum_{{\bf r}}T^\dagger({\bf r})\mathcal{H}_\text{aux}({\bf r}_d)T({\bf r}) - N\mathcal{H}_\text{cbath},
\end{aligned}\end{equation}
where \({\bf r}\) sums over all lattice sites, and $T({\bf r})$ is a many-body translation operator that translates all 
%electrons 
coordinates by a vector \({\bf r}\). Further details of the tiling procedure are provided in the Supplemental Material~\cite{suppmat}.
%We have removed extra copies of conduction bath kinetic energy \(\mathcal{H}_\text{cbath}\) that were added during the translation operations. 
Implementing this procedure for the impurity model defined above (eq.~\ref{impurityModel}) generates a {\it Hubbard-Heisenberg} lattice model in two spatial dimensions:
\begin{equation}\begin{aligned}
		\mathcal{H}_\text{tiled} &= -\frac{1}{\sqrt\mathcal{Z}}\tilde t\sum_{\left<{\bf r}_i, {\bf r}_j\right>;\sigma}\left(c^\dagger_{{\bf r}_i,\sigma}c_{{\bf r}_j,\sigma} + \text{h.c.}\right) - \tilde \mu \sum_{{\bf r}}\hat n_{{\bf r},\sigma} \\
        &+ \frac{1}{\mathcal{Z}}\tilde J\sum_{\left< {\bf r}_i, {\bf r}_j\right>}{\bf S}_{{\bf r}_i}\cdot{\bf S}_{{\bf r}_j} - \frac{1}{2}\tilde U\sum_{\bf r}\left(\hat n_{{\bf r} , \uparrow} - \hat n_{{\bf r} , \downarrow}\right)^2  ~,
\end{aligned}\end{equation}
%where the tilde symbol indicates that the 
with parameters for the lattice model that are related to those in $\mathcal{H}_\text{aux}$
%the auxiliary model
~\cite{suppmat}. 
%By comparing the tiled model and the general lattice model, the lattice model parameters and the auxiliary model parameters can be mapped to one another. 
The form of the eigenstates of $H_{\text{tiled}}$
%the lattice Hamiltonians obtained using eq.~\ref{tilingPrescription} 
are dictated by a many-body version of Bloch's theorem~\cite{stoyanova}. Further, the relation between Hamiltonians and eigenstates of the auxiliary and tiled models can be exploited in obtaining equal-time correlation functions and
%from the auxiliary model to the lattice model, as well as equal-time correlations and 
entanglement measures of the latter from the former~\cite{suppmat}.
%such entanglement entropy and mutual information.
\begin{figure}
    \centering
    \includegraphics[width=0.49\linewidth]{SF-3.pdf}
    \includegraphics[width=0.49\linewidth]{kspaceDOS-3.pdf}
    \caption{Left: $k-$space resolved spin-spin correlation $\chi_s(d,{\bf k}) = \braket{{\bf S}_d\cdot{\bf S}_{\bf k}}$, in the peudogap phase, confirmed from the fact that only a part of the Fermi surface screens the impurity, the antinodes being the first to decouple from the impurity. This leads, through our tiling relations, to a $k-$space resolved gap on the lattice model, as seen from the DOS (right panel). The region around the antinode of the lattice Fermi surface gets replaced by a Luttinger surface of zeros.}
    \label{spinCorr}
\end{figure}

\paragraph*{Pseudogapping transition from Kondo breakdown.} To obtain various low-energy phases of the impurity model, we perform a scaling analysis of the associated Hamiltonian using the recently developed 
unitary renormalisation group (URG) method \cite{anirbanurg1}. The URG
%method 
proceeds by resolving quantum fluctuations at $T=0$ involving high-energy degrees of freedom in systems of correlated fermions~\cite{santanukagome,1dhubjhep,anirbanmott2,anirbanmott1,siddharthacpi,anirban_kondo,Patra_2023,anirbanurg2}, leading to a low-energy Hamiltonian with renormalised couplings and emergent degrees of freedom. 
%The method has been applied successfully to a variety of problems involving correlated fermions~\cite{santanukagome,1dhubjhep,anirbanmott2,anirbanmott1,siddharthacpi,anirban_kondo,Patra_2023,anirbanurg1,anirbanurg2}; 
We present the results for the present impurity model here, and the details of the calculation are presented in \cite{suppmat}. 
%We find that 
The renormalisation in the %momentum
${\bf k}$-resolved Kondo coupling \(J^{(j)}_{{\bf k}_1, {\bf k}_2}\) at the \(j^\text{th}\) RG step is obtained as
\begin{equation}\begin{aligned}\label{KondoRGequation}
	\Delta J^{(j)}_{{\bf k}_1, {\bf k}_2} = -\sum_{{\bf q} \in \text{PS}} \frac{J^{(j)}_{{\bf k}_2,{\bf q}} J^{(j)}_{{\bf q},{\bf k}_1} + 4J^{(j)}_{{\bf q}, {\bf \bar q}} W_{{\bf \bar q}, {\bf k}_2, {\bf k}_1, {\bf q}}}{\omega - \frac{1}{2}|\varepsilon_j| + J^{(j)}_{{\bf q}}/4 + W_{{\bf q}}/2}~,
\end{aligned}\end{equation}
where \(\varepsilon_j\) is the energy of the shell being decoupled at the \(j^\text{th}\) step, the sum is over all momentum states \({\bf q}\) in the particle sector (PS) of the energy shell \(\varepsilon_j\) (i.e., all states 
%that are 
occupied at \(T=0\) and in the absence of any quantum fluctuations), and  \({\bf \bar q} = {\bf q} + {\boldsymbol \pi}\). The bath interaction coupling $W_{{\bf \bar q}, {\bf k}_2, {\bf k}_1, {\bf q}}$ is found to be marginal under these transformations. While the complexity of the RG equation for $J_{{\bf k}_1, {\bf k}_2}$ (eq.\eqref{KondoRGequation}) is clarified through a detailed numerical evaluation, the structure of the equation is qualitatively similar to its counterpart in a study of the Mott MIT in infinite dimensions~\cite{Mukherjee_2023}. This permits the broad conclusion that for the case of attractive bath interactions $(W<0)$, the frustration of Kondo screening due to charge fluctuations (arising from $W$) lead to the Mott metal-insulator transition.

%As we 
Upon tuning the ratio of the bath and Kondo interactions ($W/J$) from zero to negative values, we observe that the following phases are emergent in the impurity model from the competition between the competing $J$ and $W$ couplings in eq.~\eqref{KondoRGequation}- (i) for $W/J<(W/J)_{\text{PG}}$, a local Fermi liquid phase (where the entirety of the Fermi surface participates in Kondo screening), (ii) for $\frac{W}{J} \in [(\frac{W}{J})_{\text{PG}}, (\frac{W}{J})_c]$, a local pseudogapped phase where disconnected parts of the Fermi surface around the node participate in Kondo screening, and (iii) a local moment phase for $\frac{W}{J} > (\frac{W}{J})_c$, where the impurity remains unscreened at low-energies. These can be visualised from $k-$space resolved spin correlations $\chi_s(d,{\bf k}) = \braket{{\bf S}_d\cdot{\bf S}_{\bf k}}$; see Fig.~\ref{spinCorr} for $\chi_s(d,{\bf k})$ in the PG. The values $(W/J)_{\text{PG}}$ and $(W/J)_c$ are hence defined as the entry into and exit from
%boundaries of 
the pseudogap phase. %Using our 
Upon mapping to the lattice model via tiling, we %consequently 
observe that the $T=0$ Mott transition 
%in $k-$space for 
of the 2D Hubbard-Heisenberg model proceeds from Fermi liquid to Mott insulator through an intervening pseudogap phase
%. obtain an $k$-space pseudogapped phase 
with a partially gapped Fermi surface (i.e., a vanishing electronic density of states around the antinodal regions, Fig.~\ref{spinCorr}, right panel). 
%This constitutes one of the main results of our work, and 
We now lay out the mechanism leading to the formation of the pseudogap. 
%it shows that the breakdown of Kondo screening on the 
%a realistic 2D square lattice very naturally gives rise to a pseudogapped phase with Fermi arc-like behaviour. 
%The local Fermi liquid and local moment phases for the impurity model correspond, respectively, to Fermi liquid and Mott insulating phases on the lattice.

>>>>

\documentclass[%
 reprint,
superscriptaddress,
groupedaddress,
%unsortedaddress,
%runinaddress,
%frontmatterverbose, 
%preprint,
%preprintnumbers,
%nofootinbib,
%nobibnotes,
%bibnotes,
 amsmath,amssymb,
 aps,
prl,superscriptaddress
%prb,
%rmp,
%prstab,
%prstper,
%floatfix,
]{revtex4-2}

@@

%\title{Pseudogapped non-Fermi liquid phase arising from Kondo breakdown at the Mott transition}

%\title{Long range entangled Fermi arcs in a Pseudogapped Mott transition}

\title{Pseudogapped Mott criticality: Stretching Kondo Screening to Breaking Point}

@@

\author{Abhirup Mukherjee}
\affiliation{%
 Department of Physical Sciences, Indian Institute of Science Education and Research Kolkata, Nadia - 741246, India
}
\author{S. R. Hassan}
\affiliation{The Institute of Mathematical Sciences, C.I.T. Campus, Chennai 600 113, India}
\author{Anamitra Mukherjee}
\affiliation{1 School of Physical Sciences, National Institute of Science, Education and Research, HBNI, Jatni 752050, India}
\author{N. S. Vidhyadhiraja}
\affiliation{Theoretical Sciences Unit, Jawaharlal Nehru Center for Advanced Scientific Research, Jakkur, Bengaluru 560064, India}
\author{A. Taraphder}
\affiliation{Department of Physics, Indian Institute of Technology Kharagpur, Kharagpur 721302, India}
%
\author{Siddhartha Lal}
\affiliation{%
 Department of Physical Sciences, Indian Institute of Science Education and Research Kolkata, Nadia - 741246, India
}

\date{\today}
\begin{abstract}
We present an auxiliary model approach that illuminates the centrality of the pseudogap phenomenon in the Mott transition of the Hubbard-Heisenberg model on the half-filled square lattice. Passage from Fermi liquid metal to Mott insulator involves two stages: the first involves a systematic unraveling of Kondo screening due to frustrating charge fluctuations in the conduction bath of the underlying lattice-embedded quantum impurity model. The destabilises the bulk Fermi liquid, culminating in disconnection of the Fermi surface by gapping the antinodes. In the second stage, an emergent pseudogap phase arises from an effective two-channel Kondo problem in the underlying impurity model. The resulting momentum-space Fermi arcs possess non-Fermi liquid excitations proximate to a critical Fermi surface. Upon approaching the transition, the nature of this exotic metal evolves with shrinking size of the arcs towards a singular nodal non-Fermi liquid. This accompanies the onset of long-ranged real-space entanglement and spin-flip correlations, and reveals the Mott transition on the square lattice to be well beyond the paradigm of local quantum criticality.
\end{abstract}

@@

\par\noindent\textit{Introduction.}
The origin of the pseudogap and strange metal phases of the cuprates continue to be hotly debated topics in the context of high-temperature superconductivity~\cite{keimer2015quantum}. The anomalous pseudogap (PG) phase, observed most prominently in the cuprates, exhibits nodal-antinodal dichotomy with spectral weight 
concentrated on Fermi arcs around the nodal regions~\cite{loeserKapitulnik1996,Norman1998,Hashimoto2014}. While several methods have observed a PG in the two dimensional Hubbard model ~\cite{KyungKotliar2006,MacridinAzevedo2006,WuFerrero2018,anirbanmott2,HilleAndergassen2020}, there is no general consensus regarding various aspects of the PG phase, including its relation to proximate Mott insulating and superconducting phases~\cite{FradkinRevModPhys2015,anirbanmott2,Kitatani2023,Sorella2023}, its evolution from weak- to strong-coupling ~\cite{HuangDevereaux2018,Fedor2022,SimkovicFerrero2024}, and whether nodal-antinodal dichotomy is intrinsic to the model~\cite{Hashimoto2014,Schafer2021}. Importantly, the connection between finite-temperature crossovers and zero-temperature ground states observed in various analyses remains unclear~\cite{White1998,Ido2018,ProustTaillefer2019,Ponsioen2019,Shengtao2021,XuZhang2022}.

We present here a new auxiliary model approach 
to the strong coupling physics of the particle-hole symmetric 2D Hubbard-Heisenberg model at zero temperature, revealing the origin and nature of its PG phase and subsequent Mott transition. The first step 
requires identifying an appropriate lattice-embedded quantum impurity model, and conducting a scaling analysis of its low-energy phases. Subsequently, the lattice model is  constructed by applying many-body lattice translation operators~\cite{stoyanova} on the impurity model (referred to henceforth as \textit{tiling}), obtaining thereby various quantities
of the former from the latter. Translation invariance is thus restored differently within our approach from the dynamical mean-field approximation~\cite{georges1996} and its cluster variants~\cite{Hettler2000DCA,Kotliar2001Cellular,Maier2005,Sakai2023}.

Combining the scaling analysis of the impurity model with the tiling procedure provides the momentum-space resolution crucial to unveiling the PG phenomenon, and clarifies the role of the Kondo breakdown process at the heart of the Mott transition~\cite{fabrizio2017kondophysicsmotttransition,RADEMAKER2025}. Emerging from a systematic unraveling of Kondo screening, the PG phase is separated from Fermi liquid and Mott insulating phases by correlation-driven Lifshitz transitions of the Fermi surface~\cite{sakai2009evolution}. Passage through the PG involves dynamical transfer of spectral weight between gapped anti-nodal Luttinger surfaces (zeros of the single-particle Greens function) and gapless nodal arcs with non-Fermi liquid excitations. We also observe the onset of long-ranged real-space entanglement and spin-flip correlations close to the Mott transition, rendering it beyond the paradigm of local quantum criticality~\cite{Si2001}.

\par\noindent\textit{A lattice-embedded impurity model.}
Building on insight that Kondo breakdown in a simple extension of the Anderson impurity~\cite{Mukherjee_2023} captures the physics of the Mott transition in infinite dimensions~\cite{georges1996}), we study the Kondo physics of a similar Anderson impurity model embedded within a square lattice 
\begin{equation}\begin{aligned}\label{impurityModel}
	\mathcal{H}_\text{aux} = H_\text{imp} + H_\text{imp-cbath} + H_\text{cbath}~, 
\end{aligned}\end{equation}
where $H_{imp} = - \frac{U}{2}\left(\hat n_{d \uparrow} - \hat n_{d \downarrow} \right) ^2$ is the Hamiltonian for the localised impurity site, and the impurity ($c_{d\sigma}$, with spin-1/2 moment ${\bf S}_d$) 
interacts with four nearest neighbour sites ($c_{Z\sigma}$, labelled with index $Z$) 
through local Kondo terms of uniform strength $J$ as well as single-particle hybridisation $V$, \(H_\text{imp-cbath} =\frac{1}{2} J\sum_{\sigma_1,\sigma_2}\sum_{Z} {\bf S}_d\cdot c^\dagger_{Z\sigma_1}{\boldsymbol \tau}_{\sigma_1,\sigma_2} c_{Z\sigma_2} - V \sum_{\sigma, Z} \left(c^\dagger_{Z\sigma} c_{d\sigma} + h.c.\right)\). Finally, the term
$H_\text{cbath} = \sum_{{\bf k}}\epsilon_{\textbf{k}}
c^\dagger_{{\bf k},\sigma}c_{{\bf k},\sigma} -\frac{W}{2}\sum_{Z} \left(n_{Z\uparrow} - n_{Z\downarrow}\right)^2$ includes the kinetic energy ($\epsilon_{\textbf{k}}=-2t\left(\cos k_x + \cos k_y\right)$) for conduction bath tight-binding electrons ($c^\dagger_{{\bf k},\sigma}$) 
on a half-filled square lattice (with lattice spacing set to unity), together with local correlations ($W$) that frustrate the Kondo screening. The 2D structure of the impurity model, and its distinction from the infinite-dimensional counterpart, becomes apparent upon Fourier transforming the Kondo coupling to ${\bf k}$-space: 
$J_{{\bf k}, {\bf k}^\prime} = \frac{J}{2}\left[\cos\left({\bf k}_x - {\bf k}^\prime_x\right) + \cos\left({\bf k}_y - {\bf k}^\prime_y\right)\right]$, with $J>0$. Clearly, $J_{{\bf k}, {\bf k}^\prime}$ 
possesses the $C_{4}$ lattice symmetry. 
\par\noindent\textit{Reconstructing the lattice model.}
We now define the {\it tiling} procedure by which we recreate the lattice model from the impurity model Hamiltonian. For a {\it unit cell} of the tiling,  
we consider an impurity model \(\mathcal{H}_\text{aux}({\bf r}_d)\) that has the impurity site at a reference site \({\bf r}_d\) of our lattice. In order to create the lattice model, we translate the unit cell across all sites of the lattice:
\begin{equation}\begin{aligned}
	\label{tilingPrescription}
	\mathcal{H}_\text{tiled} &= \sum_{{\bf r}}T^\dagger({\bf r})\mathcal{H}_\text{aux}({\bf r}_d)T({\bf r}) - N\mathcal{H}_\text{cbath},
\end{aligned}\end{equation}
where \({\bf r}\) sums over all lattice sites, and $T({\bf r})$ is a many-body translation operator that translates all coordinates by a vector \({\bf r}\). Further details of the tiling procedure are provided in the Supplemental Material~\cite{suppmat}. Implementing this procedure for the extended Anderson impurity model defined above (eq.~\ref{impurityModel}) generates a {\it Hubbard-Heisenberg} lattice model in two spatial dimensions:
    \begin{eqnarray}
        &&\mathcal{H}_\text{tiled} = -\frac{\tilde t}{\sqrt\mathcal{Z}}\sum_{\left<{\bf r}_i, {\bf r}_j\right>;\sigma}\left(c^\dagger_{{\bf r}_i,\sigma}c_{{\bf r}_j,\sigma} + \text{h.c.}\right) - \tilde \mu \sum_{{\bf r}}\hat n_{{\bf r},\sigma}\nonumber \\
        &&+ \frac{\tilde J}{\mathcal{Z}}\sum_{\left< {\bf r}_i, {\bf r}_j\right>}{\bf S}_{{\bf r}_i}\cdot{\bf S}_{{\bf r}_j} - \frac{1}{2}\tilde U\sum_{\bf r}\left(\hat n_{{\bf r} , \uparrow} - \hat n_{{\bf r} , \downarrow}\right)^2~,
        \end{eqnarray}
with parameters $(\tilde{t},\tilde{\mu},\tilde{J},\tilde{U})$
that are related simply to those in $\mathcal{H}_\text{aux}$~\cite{suppmat}. The form of the eigenstates of $H_{\text{tiled}}$ are dictated by a many-body version of Bloch's theorem~\cite{stoyanova}. Further, relations between the Hamiltonians and eigenstates of the impurity and lattice models are exploited in obtaining equal-time correlation functions and entanglement measures of the latter from the former~\cite{suppmat}.
\begin{figure}
    \centering
    \includegraphics[width=0.31\linewidth, height=2.45cm]{phaseDiagram.pdf}
    \includegraphics[width=0.67\linewidth]{SF-DOS.pdf}
    \caption{Left: Phase diagram of impurity model at strong coupling in $U$ in terms of competing dimensionless Kondo ($J/t$) and bath correlation ($W/t$) couplings. A pseudogap phase (red, PG) is observed between the local Fermi liquid (pink, LFL) and local moment (black, LM) phases. Center: $k-$space resolved spin-spin correlation $\chi_s(d,{\bf k}) = \braket{{\bf S}_d\cdot{\bf S}_{\bf k}}$ in the pseudogap phase of the impurity model. Antinodal regions are observed to decouple from Kondo screening of the impurity. Right: Upon tiling, this leads to a $k-$space resolved antinodal gap in the electronic density of states of the lattice model, corresponding to Luttinger surfaces of zeros.}
    \label{spinCorr}
\end{figure}
\par\noindent\textit{Pseudogapping transition from Kondo breakdown.} A detailed picture of Kondo breakdown in the large $U$ phase of $H_\text{aux}$ is obtained from a scaling analysis of the ${\bf k}$-resolved Kondo coupling \(J^{(j)}_{{\bf k}_1, {\bf k}_2}\) using the 
unitary renormalisation group (URG) method \cite{anirbanurg1} (see \cite{suppmat} for details)
\begin{equation}\begin{aligned}\label{KondoRGequation}
	\Delta J^{(j)}_{{\bf k}_1, {\bf k}_2} = -\sum_{{\bf q} \in \text{PS}} \frac{J^{(j)}_{{\bf k}_2,{\bf q}} J^{(j)}_{{\bf q},{\bf k}_1} + 4J^{(j)}_{{\bf q}, {\bf \bar q}} W_{{\bf \bar q}, {\bf k}_2, {\bf k}_1, {\bf q}}}{\omega - \frac{1}{2}|\varepsilon_j| + J^{(j)}_{{\bf q}}/4 + W_{{\bf q}}/2}~,
\end{aligned}\end{equation}
where \(\varepsilon_j\) is the energy of the shell being decoupled at the \(j^\text{th}\) step, the sum is over all momentum states \({\bf q}\) in the particle sector (PS) of the energy shell \(\varepsilon_j\) (i.e., all states occupied at \(T=0\) and in the absence of any quantum fluctuations), and  \({\bf \bar q} = {\bf q} + {\boldsymbol \pi}\) is the particle-hole transformed state associated with ${\bf q}$. The bath interaction coupling $W_{{\bf \bar q}, {\bf k}_2, {\bf k}_1, {\bf q}}$ is found to be marginal under these transformations. While the complexity of the RG equation for $J_{{\bf k}_1, {\bf k}_2}$ (eq.\eqref{KondoRGequation}) is clarified through a detailed numerical evaluation whose results are discussed below, its structure permits the broad conclusion that the frustration of Kondo screening due to charge fluctuations (arising for attractive bath interactions $W<0$) lead to the Mott metal-insulator transition~\cite{Mukherjee_2023}.

Upon tuning the ratio of the bath and Kondo interactions ($W/J$) from zero to negative values (see phase diagram in Fig.\ref{spinCorr}(left)), we observe that the following phases are emergent in the impurity model from the competition between the competing $J$ and $W$ couplings in eq.~\eqref{KondoRGequation}- (i) for $W/J<(W/J)_{\text{PG}}$, a local Fermi liquid phase (where the entirety of the Fermi surface participates in Kondo screening), (ii) for $\frac{W}{J} \in [(\frac{W}{J})_{\text{PG}}, (\frac{W}{J})_c]$, a local PG phase where disconnected parts of the Fermi surface around the node participate in Kondo screening, and (iii) a local moment phase for $\frac{W}{J} > (\frac{W}{J})_c$, where the impurity remains unscreened at low-energies. These can be visualised from $k-$space resolved spin correlations $\chi_s(d,{\bf k}) = \braket{{\bf S}_d\cdot{\bf S}_{\bf k}}$; see Fig.\ref{spinCorr}(center) for $\chi_s(d,{\bf k})$ in the PG. The values $(W/J)_{\text{PG}}$ and $(W/J)_c$ are hence defined as the entry into and exit from the PG phase. Upon mapping to the lattice model via tiling, we observe that the $T=0$ Mott transition of the 2D Hubbard-Heisenberg model proceeds from Fermi liquid to Mott insulator through an intervening PG phase with a partially gapped Fermi surface (i.e., a vanishing electronic density of states around the antinodal regions, Fig.\ref{spinCorr}(right)). We now lay out the mechanism for the formation of the PG.

\begin{figure}[htpb]
    \centering
    \includegraphics[width=\linewidth]{zerosFlow.pdf}
    \caption{Left, Right and Center: Decoupling of $J_{{\bf k}, {\bf k'}}$ (positive, negative and zeros shown in yellow, purple and green respectively), with ${\bf k}$ (grey circle) corresponding to the node, antinode and a point mid-way between them on the top right arm of the Fermi surface respectively. This is seen via the appearance of zeros (green patches) of $J_{{\bf k}, {\bf k'}}$ for ${\bf k'}$ initially on the nodal regions of adjacent arms, and their subsequent progression towards the antinodes. The decoupling 
    ends with the onset of the pseudogap.
    }
    \label{rgProgression}
\end{figure}
